Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-02-10 Paper/Draft_harmonization_paper.tex
10.02.: Small changes

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. 
%Version from Brooke: As a basis for evaluating the comparability of freshwater invertebrate databases, we developed four novel invertebrate trait datasets by harmonising seven grouping features from existing trait databases. We then evaluated the influence of different...
Identified trait definitions discrepancies served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features from existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation approaches on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of taxonomic hierarchy and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

\begin{figure}
  \centering
  \subfile{Flowchart/Flowchart_methods.tex}
  \caption{Data processing steps of the selected traits. Intermediate (gray) and main (orange) steps of data preparation are depicted. The dashed bottom box illustrates the different trait aggregation methods (hypothetical data in the upper left corner). The aggregation methods (purple) and intermediate steps of the aggregation methods (pink) are displayed. Abbreviations: EU, Europe; NOA, North America; AUS, Australia; NZ, New Zealand.}
  \label{fig:data_proc_overview}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for Australia and North America for a subset of grouping features and taxa. For the Australian dataset, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from \citet{chessman_dissolved-oxygen_2018}. In \citet{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the Australian dataset except that the trait parasite is missing. We conducted the comparison for the 220 families available in \citet{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounted to 1760 cases.

For the North American dataset, we compared aggregated trait affinities with assigned trait affinities on family-level for the grouping features feeding mode, respiration, size, voltinism, and locomotion. The assigned trait affinities at family-level are part of the CONUS database and originate from expert knowledge. Trait information was available for 94 families of which all were present in the North American dataset (total number of cases 1598). The traits were on the categorical scale and were converted to binary traits prior to the comparison with aggregated trait affinities.

As mentioned above, trait affinities ranged from 0 to 1. Hence, the maximum difference possible in trait affinities is 1 or -1 (corresponds to 100 \%). For convenience and to improve interpretation, we report absolute trait differences.

@@

As ecological knowledge increases, and with the aid of new  methods such as computer vision for species and trait identification (\cite{hoye_deep_2020}), more observational and species-level trait data at may become available in the future, reducing the need for trait aggregation. However, we showed that trait aggregation is often in agreement with expert assignments, especially through aggregation approaches that use the median.

No aggregation method is objectively best, and except for cases where traits are highly variable and taxonomic hierarchical structure is highly uneven, choice of aggregation method does not appear to be influential on the outcome. Weighting trait information based on the taxa present in the database used does not seem to have much influence except when the taxonomic hierarchical structure is highly uneven. However, we recommend that the type of coding used to describe traits (binary, fuzzy) should be considered, because median and mean aggregations can yield very different results for binary-coded traits. We also suggest that a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. Through the use of carefully synthesised data from freshwater invertebrate trait databases, while acknowledging the limitations inherent in harmonised and aggregated data, the ability to investigate a wide range of large-scale ecological questions is enhanced.

>>>>

%Version from Brooke: As a basis for evaluating the comparability of freshwater invertebrate databases, we developed four novel invertebrate trait datasets by harmonising seven grouping features from existing trait databases. We then evaluated the influence of different...

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Identified trait definitions discrepancies served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features from existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation approaches on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of taxonomic hierarchy and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

\begin{figure}
  \centering
  \subfile{Flowchart/Flowchart_methods.tex}
  \caption{Data processing steps of the selected traits. Intermediate (grey) and main (orange) steps of data preparation are depicted. The dashed bottom box illustrates the different trait aggregation methods (hypothetical data in the upper left corner). The aggregation methods (purple) and intermediate steps of the aggregation methods (pink) are displayed. Abbreviations: EU, Europe; NOA, North America; AUS, Australia; NZ, New Zealand.}
  \label{fig:data_proc_overview}
\end{figure}

@@

@@

As ecological knowledge increases, and with the aid of new  methods such as computer vision for species and trait identification (\cite{hoye_deep_2020}), more observational and species-level trait data at may become available in the future, reducing the need for trait aggregation. However, we showed that trait aggregation is often in agreement with expert assignments, especially through aggregation approaches that use the median. No aggregation method is objectively best, and except for cases where traits are highly variable and taxonomic hierarchical structure is highly uneven, choice of aggregation method does not appear to be influential on the outcome. Weighting trait information based on the taxa present in the database used does not seem to have much influence except when the taxonomic hierarchical structure is highly uneven. However, we recommend that the type of coding used to describe traits (binary, fuzzy) should be considered, because median and mean aggregations can yield very different results for binary-coded traits. We also suggest that a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. Through the use of carefully synthesised data from freshwater invertebrate trait databases, while acknowledging the limitations inherent in harmonised and aggregated data, the ability to investigate a wide range of large-scale ecological questions is enhanced.
